Data organization in spreadsheets

### 2017-03-15 manuscript.Rmd
A number of minor changes (typos/style, for the most part)
Also:

 - added Lincoln Mullen to acknowledgmets

 - moved text about multiple worksheets in an Excel file to earlier in
   that section

On this point, [@ziemann2016] studied gene lists contained within the
supplementary files from 18 journals for the years 2005-2015, and
found that ~20% of gene lists had errors in the gene names, related to
the conversion of gene symbols to dates or floating-point numbers.

@@

Figure 2 contains two examples of spreadsheets with some empty cells.
In Figure 2A, cells where left blank when a single value was meant to be
repeated multiple times. Please don't do this!
It is additional work for the analyst to determine the implicit values for these cells.
Moreover, if the rows were sorted at some point, the
date column could be completely mangled.

@@

You could fill in some of those cells, to make it more clear, but even
better would be to make a "tidy" version of the data [@wickham2014],
with each row being one replicate and the response values all in one column, as in
Figure 3. We will discuss this in more detail, below.

@@

For example, you might have a column with "`plate position`" as
"`plate-well`", such as "`13-A01`". It would be better to separate this
into "`plate`" and "`well`" (containing "`13`" and "`A01`"), or even
"`plate`", "`well_row`", and "`well_column`" (containing "`13`", "`A`",
and "`1`").

@@

Another common situation is to include a note within a cell, with the
data, like "`0 (below threshold)`". Instead, write "`0`" and include a
separate column with notes.

@@

Some data sets won't fit nicely into a single rectangle, but they will usually
fit into a set of rectangles, in which case you can make a set of Excel files,
each with a rectangle of data. It is best to keep each rectangle in its own
file; tables scattered around a worksheet are difficult to work with, and they
make it hard to export the data to CSV files (which we will discuss shortly).

@@

Have sympathy for your analyst (which could be yourself): organize your data as
a rectangle (or, if necessary, as a set of rectangles). This raises the question
of whether to use multiple worksheets within one Excel file, or whether it would
be better to use multiple files, each with one worksheet. We prefer to have
multiple files with one sheet each so we can save the data as CSV files (more on
this later), but if you do use multiple worksheets in a file be sure to use a
consistent structure.

@@

An example data dictionary is displayed in Figure 9.
Note that this is a rectangular data set, like any other. The first
column contains the variable names. The second column is more readable
version, as might be used in data visualizations. The third column
groups the variables into different categories, which might
also be used in data visualizations. The last column is a description.

@@

As a general rule, don't use spaces, either in variable names (that
is, the names of the columns in your data) or in
file names. They make programming harder: the analyst will need to
surround everything in double quotes, like `"glucose 6 weeks"` rather
than just writing `glucose_6_weeks`. Where you might use spaces, use
underscores or perhaps hyphens. But don't use a mixture of underscores
and hyphens; pick one and be consistent.

@@

Keep all versions of the data files, so that if something gets
corrupted (e.g., you accidentally type over some of the data without
noticing it until much later), you will be able to go back and fix it.
Before you start inserting more data, make a copy of the file with a
new version number: `file_v1.xlsx`, `file_v2.xlsx`, ...

When you are not actively entering data, and particularly when you are
done entering data, **write-protect the file**. That way, you won't
accidentally change things.

@@

**Back up your data!**

@@

Lance Waller generously provided comments to improve the manuscript.

>>>>

On this point, Ziemann et al. [-@ziemann2016] studied gene lists contained within the
supplementary files from 18 journals for the years 2005-2015, and
found that ~20% of gene lists had errors in the gene names, related to
the conversion of gene symbols to dates or floating-point numbers.

@@

Figure 2 contains two examples of spreadsheets with some empty cells.
In Figure 2A, cells were left blank when a single value was meant to be
repeated multiple times. Please don't do this!
It is additional work for the analyst to determine the implicit values for these cells.
Moreover, if the rows are sorted at some point, the
date column will be completely mangled.

@@

You could fill in some of those cells, to make it more clear, but even
better would be to make a "tidy" version of the data [@wickham2014],
with each row being one replicate and with the response values all in one column, as in
Figure 3. We will discuss this in more detail, below.

@@

For example, you might have a column with "`plate position`" as
"`plate-well`", such as "`13-A01`". It would be better to separate this
into "`plate`" and "`well`" columns (containing "`13`" and "`A01`"), or even
"`plate`", "`well_row`", and "`well_column`" (containing "`13`", "`A`",
and "`1`").

@@

Another common situation is to include a note within a cell, with the
data, like "`0 (below threshold)`". Instead, write "`0`" and include a
separate column with such notes.

@@

Some data sets won't fit nicely into a single rectangle, but they will usually
fit into a set of rectangles, in which case you can make a set of Excel files,
each with a rectangle of data. It is best to keep each rectangle in its own
file; tables scattered around a worksheet are difficult to work with, and they
make it hard to export the data to CSV files (which we will discuss shortly).
You might also consider having a single Excel file with multiple
worksheets. We prefer to have multiple files with one sheet each so we
can more easily save the data as CSV files, but if you do use multiple
worksheets in a file be sure to use a consistent structure.

@@

Have sympathy for your analyst (which could be yourself): organize your data as
a rectangle (or, if necessary, as a set of rectangles).

@@

An example data dictionary is displayed in Figure 9.
Note that this is a rectangular data set, like any other. The first
column contains the variable names. The second column is a more readable
version, as might be used in data visualizations. The third column
groups the variables into different categories, which might
also be used in data visualizations. The last column is a description.

@@

As a general rule, don't use spaces, either in variable names (that
is, the names of the columns in your data) or in
file names. They make programming harder: the analyst will need to
surround everything in double quotes, like `"glucose 6 weeks"`, rather
than just writing `glucose_6_weeks`. Where you might use spaces, use
underscores or perhaps hyphens. But don't use a mixture of underscores
and hyphens; pick one and be consistent.

@@

Keep all versions of the data files, so that if something gets
corrupted (e.g., you accidentally type over some of the data and don't
notice it until much later), you will be able to go back and fix it.
Before you start inserting more data, make a copy of the file with a
new version number: `file_v1.xlsx`, `file_v2.xlsx`, ...

When you are not actively entering data, and particularly when you are
done entering data, *write-protect the file*. That way, you won't
accidentally change things.

@@

*Back up your data!*

@@

Lance Waller and Lincoln Mullen generously provided comments to improve the manuscript.
